# Working with missing data

There exists a specialized package for just about everything. One package that is pretty handy is [naniar](https://naniar.njtierney.com/) [@tierney2023]. It has functionality for finding missing values, visualizing where the missings are in a dataset (per variable or per case), and for handling them.

They even wrote a whole book about it, the Missing Book [@Tierney.Horst2022], which is absolutely worth a look.

## Types of missing data

We can make a distinction between different types of missing data, see @tbl-mcar-mnar.

|            | MCAR                                   | MAR                                      | MNAR                  |
|------------------|------------------|-------------------|------------------|
| Stands for | Missing completely at random           | Missing at random                        | Missing not at random |
| Means      | No patterns in the missings whatsoever | Random, but related to the data observed | Not random            |
| Implies    | Can be safely deleted, or imputed      | Should be imputed                        | Will bias the results |

: Types of missing data {#tbl-mcar-mnar}

## Invisible missing data in longitudinal datasets

A noteworthy case is how to handle "invisible" missing data, for example in experience sampling data where participants have missed a couple of beeps that are simply not listed in the dataset. A way to handle this, is with the function `complete` from `tidyr`.

An example:

```{r}
#| warning: false
library(tidyverse)
set.seed(123)

sample_data <- expand.grid(participant=1:5, beep=1:10)
sample_data$score <- runif(50, 0, 100)

# Remove 10% of the data randomly
sample_data <- slice_sample(sample_data, prop = .7) %>% 
  arrange(participant, beep)

slice_head(sample_data, n = 15)
```

In this dataset there are no *real* missings -- every cell in the table is filled with data, but there are missing beeps.

```{r}
sample_with_explicit_missings <- sample_data %>% 
  tidyr::complete(participant, beep)

slice_head(sample_with_explicit_missings, n = 15)
```

## Visualizing missing data {#sec-visualizing-missing}

The missing scores have now become explicit and are visible as `NA` see @fig-visible-missings and @fig-missings-with-span.

```{r}
#| label: fig-visible-missings
#| fig-cap: "Missing variables per participant from the sample dataset"
#| warning: false
 
library(naniar)
gg_miss_var(sample_with_explicit_missings, facet = participant)
```

```{r}
#| label: fig-missings-with-span
#| fig-cap: "Proportion of missing values per participant"


gg_miss_span(sample_with_explicit_missings, score, span_every = 10) +
  labs(x = "Participant")
```

It is also possible to visualize the missing data in a scatterplot (look at `naniar::geom_miss_point()`). In @fig-missing-scatter it can be seen that the missing values are approximately randomly distributed across the beeps (which makes sense because I used a random sample to create the missings in the first place).

```{r}
#| label: fig-missing-scatter
#| fig-cap: "A scatterplot showing missing values as red dots"

sample_with_explicit_missings %>%
  ggplot(aes(x = beep, y = score)) +
  naniar::geom_miss_point()
```

## Data imputation

I was taught to be very careful with data imputation. However, there can be cases where imputation is a wise thing to do because imputing a couple of missings can really save a lot of data from having to be thrown away.

Imputation methods from the `naniar` package include `impute_below` and `impute_mean`. (See documentation for that package).

A package that can handle imputation with linear models is `simputation` ([simputation github](https://github.com/markvanderloo/simputation)).

## Other random thoughts

Tidyr also has a function `fill` to "fill down" missing data, i.e. fill the empty values in a column with the last non-empty value above it.

## Resources

[Datacamp course - Dealing with missing data](https://campus.datacamp.com/courses/dealing-with-missing-data-in-r)
